A numerical solver for nonlinear electrokinetic phenomena

### 2012-03-19 notes/article/main.tex
Work on LaTeX.

%\usepackage{setspace}\doublespacing

\usepackage{graphicx}
\usepackage{tikz}
\usepackage{amsmath} 
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{units}   
\usepackage{float}   
\usepackage{graphicx}
\usepackage{amsfonts}
\usepackage{mathrsfs}
\usepackage[colorlinks]{hyperref}
\usepackage{framed}

@@

The goal of this work is to implement an accurate and fast iterative numerical
solver for nonlinear electrokinetic problems, and apply it to the study
of systems that have no closed-form solutions.
It is expected to gain insight into the chemical and physical behavior 
in far more general regimes than are currently well-understood, 
and also lead to further analytical developments, based on new scaling behavior
provided by the numerical solver. We may also target interesting physical
phenomena that have been observed experimentally.

@@

Object-Oriented Design methodology is used for implementation.
The MATLAB programming language is chosen due to strong numerical capabilities
and high-level language features.

@@

In order to validate the numerical solver, an asymptotic analysis of the nonlinear system
is performed as described in \ref{sec:asymp}. The numerical results are compared to 
to the asymptotic solutions, and good correspondence is achieved.

\subsection{Linear Regime}
For $\beta \ll 1$, the linear regime takes place:
\begin{eqnarray*}
\cU = 2\log\pars{\frac{1 + \gamma^{-\frac{1}{2}}}{2}}\beta.
\end{eqnarray*}
Thus, for $\gamma > 1$, the Debye layer has positive space charge, corresponding to a negative
particle charge -- yielding negative drift velocity, $\cU < 0$. 
For $\gamma < 1$, we have $\cU > 0$, respectively.

The linear regime is simulated for $\gamma > 1$ (at Figure \ref{Linear1})  
and for $\gamma < 1$ (at Figure \ref{Linear2}).

\subsection{Cubic Regime}
When $\gamma \approx 1$, the cubic regime, $\cU = O(\beta^3)$, 
dominates the linear one, even for weak electric fields.

This phenomenon is confirmed both by the numerical
results and by asymptotic expansion of the nonlinear system (for $\alpha=0$).
The switch from the linear regime to the cubic one happens at a critical electric
field value $\beta_c = O(\sqrt{|\gamma - 1|})$.

The numerical results, as well as the analytical ones, are given at Figure
\ref{} (for $\gamma > 1$) and at Figure \ref{} (for $\gamma < 1$).
Since log-log scale is employed, positive and negative velocities are shown in
the same plot using different markers and line styles.

\subsection{Vortices}
The linear solution for $\gamma \approx 1$, 
yields a sign change in the zeta potential $\zeta = \log (C/\gamma)$, when
$|\log\gamma| \approx 3\beta / 4$. This results in a sign change in the slip velocity,
which can be approximated for small $|\zeta| \ll 1$ as:
\begin{eqnarray*}
V_\theta \approx \zeta \cdot \bnabla_S \varPhi.
\end{eqnarray*} 
This sign change creates a vortex at the downstream part of the ion-exchanger, 
as demonstrated by the numerical simulation and the asymptotic analysis results 
in Figure \ref{}.

%        \includegraphics[width=0.7\textwidth]{figs}

\begin{figure}[htbp]
    \begin{center}
    \includegraphics[width=0.45\textwidth]{figs/Linear/g=2.00.eps}
    \includegraphics[width=0.45\textwidth]{figs/Linear/g=5.00.eps}
    \includegraphics[width=0.45\textwidth]{figs/Linear/g=10.00.eps}
    \includegraphics[width=0.45\textwidth]{figs/Linear/g=100.00.eps}
    \caption{Steady-state velocity $\cU$ as a function of the 
        electric field magnitude $\beta$, for $\gamma > 1$. 
        The analytic solution velocity is represented by dashed lines, and the numerical
        results are represented by squares.}
    \label{Linear1}
    \end{center}
\end{figure}

\begin{figure}[htbp]
    \begin{center}
    \includegraphics[width=0.45\textwidth]{figs/Linear/g=0.50.eps}
    \includegraphics[width=0.45\textwidth]{figs/Linear/g=0.20.eps}
    \includegraphics[width=0.45\textwidth]{figs/Linear/g=0.10.eps}
    \includegraphics[width=0.45\textwidth]{figs/Linear/g=0.01.eps}
    \caption{Steady-state velocity $\cU$ as a function of the 
        electric field magnitude $\beta$, for $\gamma < 1$. 
        The analytic solution velocity is represented by solid lines, and the numerical
        results are represented by circles.}
    \label{Linear2}
    \end{center}
\end{figure}

\begin{figure}[htbp]
    \begin{center}
    \includegraphics[width=0.45\textwidth]{figs/Cubic/g=0.999900.eps}
    \includegraphics[width=0.45\textwidth]{figs/Cubic/g=0.999000.eps}
    \includegraphics[width=0.45\textwidth]{figs/Cubic/g=0.990000.eps}
    \includegraphics[width=0.45\textwidth]{figs/Cubic/g=0.900000.eps}
        \caption{Steady-state velocity $\cU$ as a function of the 
        electric field magnitude $\beta$, for $\gamma < 1$. 
        The analytic solution velocity is represented by solid lines, 
        and the numerical results are represented by square markers.}
    \end{center}
    \label{fig:Cubic1}
\end{figure}
\begin{figure}[htbp]
    \begin{center}
    \includegraphics[width=0.45\textwidth]{figs/Cubic/g=1.000100.eps}
    \includegraphics[width=0.45\textwidth]{figs/Cubic/g=1.001000.eps}
    \includegraphics[width=0.45\textwidth]{figs/Cubic/g=1.010000.eps}
    \includegraphics[width=0.45\textwidth]{figs/Cubic/g=1.100000.eps}
        \caption{Steady-state velocity $\cU$ as a function of the 
        electric field magnitude $\beta$, for $\gamma < 1$. 
        The analytic solution velocity is represented by lines 
        (solid lines for $\cU > 0$, dashed lines for $\cU < 0$), 
        and the numerical results are represented by markers 
        (squares for $\cU > 0$, circles for $\cU < 0$).}
    \end{center}
    \label{fig:Cubic2}
\end{figure}
\begin{figure}[htbp]
    \begin{center}
	\includegraphics[width=0.45\textwidth]{figs/Stream/N0.01.eps}
	\includegraphics[width=0.45\textwidth]{figs/Stream/A0.01.eps}
	\includegraphics[width=0.45\textwidth]{figs/Stream/N0.02.eps}
	\includegraphics[width=0.45\textwidth]{figs/Stream/A0.02.eps}
	\includegraphics[width=0.45\textwidth]{figs/Stream/N0.04.eps}
	\includegraphics[width=0.45\textwidth]{figs/Stream/A0.04.eps}
	\includegraphics[width=0.45\textwidth]{figs/Stream/N0.08.eps}
	\includegraphics[width=0.45\textwidth]{figs/Stream/A0.08.eps}
        \caption{Streamlines of the flow for various electric field $\beta$ values: 
        numerical results (left) and analytical approximation (right) for 
        $\gamma = 0.99$.  The steady-state velocity $\cU$ is positive, so the
        ion-exchanger drifts to the right.}
    \end{center}
    \label{fig:Vortex}
\end{figure}

\begin{appendix}
\section{Asymptotic Analysis} \label{sec:asymp}
The nonlinear system is defined by the following PDEs:
\label{PDEs}
\begin{eqnarray*}
\bnabla \cdot \pars{C \bnabla \varPhi} &=& 0, \\
\bnabla \cdot \pars{\bnabla C - \alpha \bV C} &=& 0, \\
\bLaplacian \bV - \bnabla P + \Laplacian \varPhi \bnabla \varPhi &=& \bzero.
\end{eqnarray*}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Linear Solution $(k=1)$}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Nonlinear Solution ($k > 1$)}
Note that, due to symmetry considerations, $\cU(\beta)$ is an anti-symmetric function:
\begin{eqnarray*}
\cU(-\beta) &=& -\cU(\beta).
\end{eqnarray*}
Therefore, $\cU_2 = 0$ and the next velocity term is the cubic one:
\begin{eqnarray*}
\cU(\beta) &\approx& \beta \cU_1 + \beta^3 \cU_3 + O(\beta^5)
\end{eqnarray*}

>>>>

\usepackage{setspace}\doublespacing

\usepackage{graphicx}
\usepackage{tikz}
\usepackage{amsmath} 
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{units}   
\usepackage{float}   
\usepackage{graphicx}
\usepackage{amsfonts}
\usepackage{mathrsfs}
\usepackage[colorlinks]{hyperref}
\usepackage{framed}
\usepackage{url}

@@

The goal of this work is to implement an iterative numerical
solver for nonlinear electrokinetic problems, and apply it to the study
of systems that have no closed-form solutions.
It is expected to gain insight into the chemical and physical behavior 
in far more general regimes than are currently well-understood, 
and also lead to further analytical developments, based on new scaling behavior
provided by the numerical solver. We may also target interesting physical
phenomena that have been observed experimentally.

@@

Object-Oriented Design methodology is used for implementation.
The MATLAB programming language is chosen due to strong numerical capabilities
and high-level language features.
The source code of solver can be found at \cite{source}.

@@

In order to validate the numerical solver, an asymptotic analysis of the nonlinear system
is performed as described in \ref{app:asymp}. The numerical results are compared to 
to the asymptotic solutions, and good correspondence is achieved.

\subsection{Linear Regime}
For $\beta \ll 1$, the linear regime takes place (for derivation, see \ref{app:linear}):
\begin{eqnarray*}
\cU = 2\log\pars{\frac{1 + \gamma^{-\frac{1}{2}}}{2}}\beta.
\end{eqnarray*}
Thus, for $\gamma > 1$, the Debye layer has positive space charge, corresponding to a negative
particle charge -- yielding negative drift velocity, $\cU < 0$. 
For $\gamma < 1$, we have $\cU > 0$, respectively.

The linear regime is simulated for $\gamma > 1$ (at Figure \ref{fig:Linear1})  
and for $\gamma < 1$ (at Figure \ref{fig:Linear2}).

\subsection{Cubic Regime}
When $\gamma \approx 1$, the cubic regime  (for derivation, see \ref{app:cubic}), where
\begin{eqnarray*}
\cU = O(\beta^3), 
\end{eqnarray*}
dominates the linear one, even for weak electric fields.

This phenomenon is confirmed both by the numerical
results and by asymptotic expansion of the nonlinear system (for $\alpha=0$).
The switch from the linear regime to the cubic one happens at a critical electric
field value
\begin{eqnarray*}
\beta_c = O(\sqrt{|\gamma - 1|}).
\end{eqnarray*}

The numerical results, as well as the analytical ones, are given at Figure
\ref{fig:Cubic1} (for $\gamma > 1$) and at Figure \ref{fig:Cubic2} (for $\gamma < 1$).
Since log-log scale is employed, positive and negative velocities are shown in
the same plot using different markers and line styles.

\subsection{Vortices}
The linear solution for $\gamma \approx 1$, 
yields a sign change in the zeta potential $\zeta = \log (C/\gamma)$, when
$|\log\gamma| \approx 3\beta / 4$. This results in a sign change in the slip velocity,
which can be approximated for small $|\zeta| \ll 1$ as:
\begin{eqnarray*}
V_\theta \approx \zeta \cdot \bnabla_S \varPhi.
\end{eqnarray*} 
This sign change creates a vortex at the downstream end of the ion-exchanger, 
as demonstrated by the numerical simulation and the asymptotic analysis results 
in Figure \ref{fig:Vortex}.

\subsection{Discussion}
The numerical solver results have good correspondence to the asymptotic linear model as
shown above.

Moreover, two interesting phenomena (cubic regime and downstream vortex) were predicted
using the numerical results, and later verified using asymptotic expansion of the
nonlinear system.

\begin{figure}[htbp]
    \begin{center}
    \includegraphics[width=0.45\textwidth]{figs/Linear/g=2.00.eps}
    \includegraphics[width=0.45\textwidth]{figs/Linear/g=5.00.eps}
    \includegraphics[width=0.45\textwidth]{figs/Linear/g=10.00.eps}
    \includegraphics[width=0.45\textwidth]{figs/Linear/g=100.00.eps}
    \caption{Steady-state velocity $\cU$ as a function of the 
        electric field magnitude $\beta$, for $\gamma > 1$. 
        The analytic solution velocity is represented by dashed lines, and the numerical
        results are represented by squares.}
    \label{fig:Linear1}
    \end{center}
\end{figure}

\begin{figure}[htbp]
    \begin{center}
    \includegraphics[width=0.45\textwidth]{figs/Linear/g=0.50.eps}
    \includegraphics[width=0.45\textwidth]{figs/Linear/g=0.20.eps}
    \includegraphics[width=0.45\textwidth]{figs/Linear/g=0.10.eps}
    \includegraphics[width=0.45\textwidth]{figs/Linear/g=0.01.eps}
    \caption{Steady-state velocity $\cU$ as a function of the 
        electric field magnitude $\beta$, for $\gamma < 1$. 
        The analytic solution velocity is represented by solid lines, and the numerical
        results are represented by circles.}
    \label{fig:Linear2}
    \end{center}
\end{figure}

\begin{figure}[htbp]
    \begin{center}
    \includegraphics[width=0.45\textwidth]{figs/Cubic/g=0.999900.eps}
    \includegraphics[width=0.45\textwidth]{figs/Cubic/g=0.999000.eps}
    \includegraphics[width=0.45\textwidth]{figs/Cubic/g=0.990000.eps}
    \includegraphics[width=0.45\textwidth]{figs/Cubic/g=0.900000.eps}
        \caption{Steady-state velocity $\cU$ as a function of the 
        electric field magnitude $\beta$, for $\gamma < 1$. 
        The analytic solution velocity is represented by solid lines, 
        and the numerical results are represented by square markers.}
	    \label{fig:Cubic1}
    \end{center}
\end{figure}
\begin{figure}[htbp]
    \begin{center}
    \includegraphics[width=0.45\textwidth]{figs/Cubic/g=1.000100.eps}
    \includegraphics[width=0.45\textwidth]{figs/Cubic/g=1.001000.eps}
    \includegraphics[width=0.45\textwidth]{figs/Cubic/g=1.010000.eps}
    \includegraphics[width=0.45\textwidth]{figs/Cubic/g=1.100000.eps}
        \caption{Steady-state velocity $\cU$ as a function of the 
        electric field magnitude $\beta$, for $\gamma < 1$. 
        The analytic solution velocity is represented by lines 
        (solid lines for $\cU > 0$, dashed lines for $\cU < 0$), 
        and the numerical results are represented by markers 
        (squares for $\cU > 0$, circles for $\cU < 0$).}
	    \label{fig:Cubic2}
    \end{center}
\end{figure}
\begin{figure}[htbp]
    \begin{center}
	\includegraphics[width=0.45\textwidth]{figs/Stream/N0.01.eps}
	\includegraphics[width=0.45\textwidth]{figs/Stream/A0.01.eps}
	\includegraphics[width=0.45\textwidth]{figs/Stream/N0.02.eps}
	\includegraphics[width=0.45\textwidth]{figs/Stream/A0.02.eps}
	\includegraphics[width=0.45\textwidth]{figs/Stream/N0.04.eps}
	\includegraphics[width=0.45\textwidth]{figs/Stream/A0.04.eps}
	\includegraphics[width=0.45\textwidth]{figs/Stream/N0.08.eps}
	\includegraphics[width=0.45\textwidth]{figs/Stream/A0.08.eps}
        \caption{Streamlines of the flow for various electric field $\beta$ values: 
        numerical results (left) and analytical approximation (right) for 
        $\gamma = 0.99$.  The steady-state velocity $\cU$ is positive, so the
        ion-exchanger drifts to the right. 
        }
    \label{fig:Vortex}
    \end{center}
\end{figure}

\begin{appendix}
\section{Asymptotic Analysis} \label{app:asymp}
The nonlinear system is defined by the following PDEs:
\label{PDEs}
\begin{eqnarray*}
\bnabla \cdot \pars{C \bnabla \varPhi} &=& 0, \\
\bnabla \cdot \pars{\bnabla C - \alpha \bV C} &=& 0, \\
\bLaplacian \bV - \bnabla P + \Laplacian \varPhi \bnabla \varPhi &=& \bzero.
\end{eqnarray*}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Linear Solution $(k=1)$} \label{app:linear}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Nonlinear Solution ($k > 1$)} \label{app:cubic}
Note that, due to symmetry considerations, $\cU(\beta)$ is an anti-symmetric function:
\begin{eqnarray*}
\cU(-\beta) &=& -\cU(\beta).
\end{eqnarray*}
Therefore, $\cU_2 = 0$ and the next velocity term is the cubic one:
\begin{eqnarray*}
\cU(\beta) &\approx& \beta \cU_1 + \beta^3 \cU_3 + O(\beta^5)
\end{eqnarray*}
